---
title: "Introduction to SAFE"
subtitle: "A step by step online tutorial"
date: "`r Sys.Date()`"
author:
  - name: Shinichi Nakagawa
  - name: Ayumi Mizuno
  - name: Coralie Williams
  - name: Santiago Ortega
  - name: Szymon M. Drobniak
  - name: Malgorzata Lagisz
  - name: Yefeng Yang
  - name: Alistair M. Senior
  - name: Daniel W. A. Noble
  - name: Erick J Lundgren
format: 
  html:
    toc: true
    toc-location: left
    toc-depth: 6
    toc-float: true
    toc-expand: true
    toc-title: "**CONTENTS**"
    output-file: "index.html"
    # embed-resources: true
    # code-fold: true
    code-link: true
    code-tools: true
    # number-sections: false
    # bibliography: ./bib/ref.bib
    fontsize: "10"
    mainfont: "Helvetica"
    monofont: "Fira Code"
    # fontcolor: ""
    page-layout: article
    code-overflow: wrap
    df_print: paged
    theme: cosmo
    code-line-numbers: false
    grid:
      margin-width: 120px
# crossref: 
#   fig-title: Figure     # (default is "Figure")
#   tbl-title: Table     # (default is "Table")
#   title-delim: "—"     # (default is ":")
#   fig-prefix: Fig.     # (default is "Figure")
#   tbl-prefix: Table.   # (default is "Table")
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
  tidy: true
bibliography: references.bib
---

## Introduction

If you have any questions, errors or bug reports, please feel free to contact Erick Lundgren (erick.lundgren\@gmail.com).

To run the code interactively (instead of just reading this lovely tutorial), do the following:

1.  Clone the github repo: https://github.com/ejlundgren/SAFE.git
2.  Run the code chunks in the SAFE simulations tutorial.qmd while in the R project session or just follow along here on the web

## Load libraries

```{r, load_libraries}
#| label: load_libraries
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

# First, clean the environment:
rm(list = ls())

# Now load packages:
library("pacman")
p_load(data.table, MASS, crayon, 
          tmvtnorm, metafor,
          parallel, foreach, doSNOW,
          ggplot2, patchwork, scico)

# Also, we'll create a theme object for ggplots to make code more readable:
theme_SAFE <- theme_bw()+
  theme(panel.border = element_blank(),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        panel.grid = element_blank(),
        strip.background = element_blank())


```

# An introduction to SAFE bootstrapping

SAFE bootstrapping is an incredible, seemingly magical, technique to calculate bias-corrected point estimates and sampling variance for meta-analytic effect sizes. SAFE bootstrapping is particularly useful for novel effect sizes for which sampling variance formulas may not be known. In some cases, it is also more unbiased than traditional plug-in formulas for variance and point estimates.

## How does it work?

With SAFE bootstrapping, one draws a cloud (or vector) of *hyperparameters* of all the variables used in an effect size formula. You then apply your effect size formula to this cloud. The variance of this transformed cloud is the sampling variance of your effect size!

You can then calculate a bias-transformed point estimate by taking the mean of this transformed cloud and subtracting it from 2 times the point estimate derived from the original variables.

Let's demonstrate.

Imagine you're conducting a meta-analysis of animal responses to a challenge. Many studies report data as mean 'latency time' (e.g., time until an animal runs or solves a task). However, you want to create an effect size of 'speed' from latency time. This is simple: take the inverse of latency (`1/latency`). Here let's start with a single mean latency time, and associated standard deviation and sample size, and convert it to our new 'speed' effect size.

```{r, speed_example}
#| label: speed_example
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"
# The type of data summaries you might extract for a meta-analysis:
mean_latency <- 15
n <- 25
sd <- 4.6

# New effect size:
1/mean_latency

```

The effect size is trivial to calculate. But what is the sampling variance of this effect size, given that each of these measurements was associated with measurement sample size and standard deviation?

You could use calculus to derive a first-order sampling variance formula. Or you could use SAFE.

### Step 1: Draw hyperparameters.

If the word 'hyperparameter' is scary, what we mean is a distribution associated with the variables in our effect size formula, in this case mean `latency`.

This is different than drawing a population with standard deviation, which will have a broad distribution, reflecting the underlying population. Instead, we're interested in drawing a distribution of the **means**, the hyperparameter of our effect size of interest. To do this, we draw 1,000,000 potential means from a normal distribution, with **standard error of the mean** to describe dispersion.

```{r, speed_cloud}
#| label: speed_cloud
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"
# Set seed for reproducibility
set.seed(2025)

# Define the variables
x <- 15   # The mean latency time
sd <- 1.6 # standard deviation of latency time
n <- 25   # sample size, e.g., number of animals in the experiment

# Simulate a vector of hyperparameters (or cloud)
cloud <- rnorm(1e6,                  # number of samples
               mean = x,             # mean latency time
               sd = sd / sqrt(n))    # standard error

# Let's have a look at the hyperparameter cloud
head(cloud)
hist(cloud)

```

### Step 2: Transform hyperparameters to target effect size

Now, let's transform this cloud of hyperparameters to 'speed' with our 'plug-in' or definition formula `1/x`. Note that when we write 'plug-in', we mean using a formula to calculate effect sizes or sampling variance.

These transformed values are centered on our original effect size, which we'll overlay in red

```{r, speed_transform}
#| label: speed_transform
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"
# Transform the cloud to our effect size
cloud_trans <- 1/cloud

# Calculate the 'plug-in' effect size of our observed mean latency time with the definition formula
x <- 15
plugin_effect_size <- 1/x
plugin_effect_size

# Compare plug-in effect size to transformed cloud of hyperparameters:
ggplot()+
  geom_density(aes(x = cloud_trans), fill = "grey80")+
  geom_vline(xintercept = plugin_effect_size, color = "red")+
  theme_SAFE

```

### Step 3: Calculate sampling variance

To calculate sampling variance, we'll simply calculate the variance of our transformed hyperparameter cloud.

```{r, speed_samplevar}
#| label: speed_samplevar
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"
safe_vi <- var(cloud_trans)

```

Voilà! Sampling variance for a novel effect size. Pretty magical right?

We can compare this to sampling variance calculated using the Delta method on our `1/x` effect size:

```{r, delta_method}
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

# Delta method sampling variance:
(sd^2 / n) * (-1 * x ^ -2)^2

# SAFE method:
safe_vi
```

We'll compare these results below (@sec-compare-results) to see whether SAFE provides as good of an estimate as this 'plug-in' estimate of sampling variance.

### Step 4: Calculate the bias-corrected effect size

It may seem that we could just take the mean of this transformed cloud of hyperparameters to obtain an unbiased estimate of our effect size. However, due to Jensen's inequality, we cannot. Jensen's inequality states that applyingOK any function (such as `1/x`) to the *mean* of a distribution does not equal taking the *mean* of that same function **applied** to the distribution.

$$
f(\bar{x}) \neq \bar{f(x)}
$$ {#eq-inequality}

We can demonstrate in `R` as follows. Here we'll compare the transformed hyperparameter cloud (`cloud_trans`) to our plug-in estimate (`1/x`) and to the mean of our hyperparameter cloud (`mean(cloud_trans)`):

```{r}
#| label: compare-estimates
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

ggplot()+
  geom_density(aes(x = cloud_trans), fill = "grey80")+
  geom_vline(xintercept = plugin_effect_size, color = "red")+
  geom_vline(xintercept = mean(cloud_trans), color = "blue")+
  theme_SAFE

```

As you can see, the mean of transformed parameter cloud (blue vertical line) is slightly higher than our plug-in effect size (red line) because the transformed cloud's distribution is somewhat skewed. The difference between these two estimates is the bias that we need to correct for to arrive at a bias-corrected point estimate.

We must therefore calculate an estimate of this bias and use this to correct our estimate. Bias measures how different the bootstrapped estimate is from the 'true value' and is calculated with the following formula:

$$
\widehat{\text{bias}} = \bar{\theta}^* - \hat{\theta}_{\text{truth}}
$$ {#eq-bias}

Where, $\bar{\theta}^*$ is the average of the bootstrapped estimates and $\hat{\theta}_{\text{truth}}$ is the true value. However, we do not know what the 'true' value is, given that our mean is a sample mean and the underlying population mean is unknown. We thus use our plug-in estimate as surrogate for 'truth'.

$$
\widehat{\text{bias}} = \bar{\theta}^* - \hat{\theta}_{\text{plugin}}
$$

This estimate of bias captures the difference between the mean of our transformed cloud of hyperparameters and the plug-in estimate, which we treat as 'truth' (also known as the 'estimand', which we will explain in more detail below).

We now use this estimate of bias to correct our plug-in estimate. Note that this is very confusing because we are treating the plug-in as both 'truth' and as an estimate that needs to be corrected.

If the bootstrapped estimate is larger than the plug-in estimate, then the plug-in estimate is biased low, and we need to push it up. If the bootstrapped estimate is smaller than the plug-in estimate, then the plug-in estimate is biased high, and we need to push it down.

Bias correction subtracts this bias to correct the point estimate. In other words, the adjustment needed depends on the magnitude of difference and sign of the bias. We can substitute the bias into the following formula for bias-corrected estimate:

$$
\hat{\theta}_{\text{bc}} = \hat{\theta}_{\text{plugin}} - \widehat{\text{bias}}
= \hat{\theta}_{\text{plugin}} - \big(\bar{\theta}^* - \hat{\theta}_{\text{plugin}}\big)
$$ {#eq-bc}

Rearranging this gives us:

$$
\hat{\theta}_{\text{bc}} = 2 \hat{\theta}_{\text{plugin}} - \bar{\theta}^*
$$

You will notice that this is the same equation we use in the code chunk below.

```{r, speed_bias_correction}
#| label: speed_bias_correction
#| eval: true 

safe_yi <- (2 * plugin_effect_size) - mean(cloud_trans)
safe_yi
```

We can now compare our plug-in effect size to our bias-corrected SAFE effect size:

```{r, speed_plugin_effect_size}
#| label: speed_plugin_effect_size
#| eval: true
plugin_effect_size

```

So similar! And potentially less biased. Let's find out using Monte Carlo simulations where we compare the bias of estimates from SAFE to plug-in estimates. We'll also demonstrate that the SAFE bias correction works by looking at the bias of the uncorrected mean of our SAFE bootstraps estimates. Note, that after these simulations we provide a guide (@sec-custom-effect-sizes) on how to use SAFE for custom effect sizes for which there may be no documented variance formula.


# Creating custom effect sizes {#sec-custom-effect-sizes}

The greatest draw of the SAFE method is its applicability to a variety of custom effect sizes that researchers may want to use for particular meta-analyses. After all, in order to fit a meta-analytic model, one requires a measure of sampling variance to use as a model weight. Deriving this sampling variance numerically is non-trivial as it requires using the Delta method and is difficult to confirm or validate. Let's try using SAFE for the temperature coefficient Q~10~.

### The Q~10~ Temperature Coefficient

Q~10~ is a widely used measure of temperature sensitivity of chemical reactions or biological processes. Q~10~ compares the rate of a process between 2 temperature treatment groups. The formula for Q~10~ is:

$$
Q_{\text10} = (R_{\text2} / R_{\text{1}})^{10/(T_{\text2}-T_{\text1})}
$$ {#eq-Q10}

Where `R` is the rate of some biological or chemical process for treatment groups 1 and 2 and `T` is the temperatures of groups 1 and 2.

If one wanted to conduct a meta-analysis on studies reporting Q~10~ data, along with sample size and standard deviation for each mean `R`, how would you calculate sampling variance? One could derive sampling variance using the delta method but this is outside many people's mathematical comfort zone and it's not clear how one would evaluate whether the derived sampling variance is accurate or not.

For example, @havird2020, conducted an influential meta-analysis using Q~10~ on passive versus active thermal acclimation among a diversity of aquatic and terrestrial taxa. At the time there was no sampling variance for Q~10~ effect sizes and so the authors derived it using the Delta method. However, SAFE could have saved the day!

Let's see if we can use SAFE to calculate sampling variance for Q~10~ type data. We will use a fake dataset, reflecting what a meta-analyst might extract from the literature.

```{r, load_Q10_data}
#| label: load_Q10_data
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

dat <- fread("data/Q10_fake_data.csv")
dat
```

This dataset has IDs for the references and observation IDs (`Reference_ID` and `Obs_ID`) as well as means, standard deviations, and sample sizes for R1 and R2 (`R1_mean`, `R2_mean`, `R1_SD`, `R2_SD`, `R1_N`, `R2_N`) and the two temperature treatments (`T1`, `T2`).

First, let's encapsulate the Q~10~ plug-in formula in a function and then calculate Q~10~ point estimates (effect sizes) for each row of our fake data.

```{r, calculate_Q10}
#| label: calculate_Q10
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

Q10 <- function(R1, R2, T1, T2){
  res <- (R2 / R1) ^ (10 / (T2 - T1))
}

dat$yi_plugin <- Q10(R1 = dat$R1_mean,
                     R2 = dat$R2_mean,
                     T1 = dat$T1,
                     T2 = dat$T2)

dat

```

### Calculate sampling variance for Q~10~

Now, let's use SAFE to calculate bias corrected point estimates and sampling variances. To do this, we will draw hyperparameter clouds for R1 and R2 based on mean `R` and standard error (`SD/sqrt(N)`), transform these clouds with our `Q10` function, calculate sampling variance of this transformed cloud, and then calculate our bias-corrected point estimates.

This has to be done separately for each row in our dataset. To make the code more readable, we'll do this in a for loop (though parallel approaches or lapply would be faster for larger datasets).

```{r, Q10_SAFE}
#| label: Q10_SAFE
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

res <- list()

for(i in 1:nrow(dat)){
  # Create clouds of hyperparameters
  R1_cloud <- rnorm(n = 1e6, 
                    mean = dat[i, ]$R1_mean,
                    sd = dat[i, ]$R1_SD / sqrt(dat[i, ]$R1_N))
  R2_cloud <- rnorm(n = 1e6, 
                    mean = dat[i, ]$R2_mean,
                    sd = dat[i, ]$R2_SD / sqrt(dat[i, ]$R2_N))
  
  # Transform clouds with Q10 formula
  Q10_cloud <- Q10(R1 = R1_cloud,
                   R2 = R2_cloud,
                   T1 = dat[i, ]$T1,
                   T2 = dat[i, ]$T2)
  
  # calculate sampling variance:
  vi_SAFE <- var(Q10_cloud)
  
  # Calculate plug-in point estimate with formula:
  plugin_effect_size <- Q10(R1 = dat[i, ]$R1_mean,
                            R2 = dat[i, ]$R2_mean,
                            T1 = dat[i, ]$T1,
                            T2 = dat[i, ]$T2)
  
  # Bias corrected SAFE estimate:
  yi_SAFE <- (2 * plugin_effect_size) - mean(Q10_cloud)
  
  # Combine data and store in list
  res[[i]] <- data.table(dat[i, ],
                         yi_SAFE = yi_SAFE,
                         vi_SAFE = vi_SAFE)
  
}
# Convert list to data.table
dat.final <- rbindlist(res)

# Print
dat.final

```

Sampling variance and point estimates for Q~10~! This sampling variance can then be used in a meta-analytic model:

```{r, Q10_meta_analysis}
#| label: Q10_meta_analysis
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

m <- rma.mv(yi_SAFE ~ 1,
            V = vi_SAFE,
            random = ~ 1 | Reference_ID/Obs_ID,
            data = dat.final)
summary(m)

```

Pretty neat, right?

### Using SAFE for custom effect sizes

If you want to use SAFE for other custom effect sizes, you'd follow a similar process:

1.  Determine which variables in your effect size formula need to be bootstrapped. Note that with Q~10~, the two variables that needed to be drawn were `R1` and `R2`, as these were associated with SD and N in the reporting studies. `T1` and `T2` were experimental constants and thus did not need to be drawn for SAFE calculations.

2.  Determine what the underlying distribution of these variables is. `R1` and `R2`, as rates, approximate the normal distribution. Other variables (e.g., counts) would need to be drawn from the binomial distribution.

3.  As long as your data consists of independent groups, you can draw your bootstrapped variables in separate `rnorm` (or `rbinom`) calls as we did above. If your data is truncated (e.g., bounded by 0 or between 0 and 1), you can draw your bootstrapped samples with the `tmvtnorm` package, where you can specify boundaries. If there is dependency between groups (i.e., paired designs), you'll need to use slightly more complicated methods with multivariate draws and sigma-matrices as described in the manuscript (Nakagawa et al. 2025). Note that multivariate draws (i.e., with `MASS::mvrnorm()`) often take variance instead of SD. In the next section, we'll share some sample code of how to use SAFE with these alternative bootstrapping functions.

After making those decisions, simply follow the 4 steps to 'SAFEty'. 🙃

## Beyond `rnorm`

While the function `rnorm` can be used for a wide variety of effect sizes, some problems may occur. For example, some effect sizes account for paired designs (e.g., before-after) and thus require multivariate bootstrapping, where the two distributions are drawn based on a covariate matrix. Others, such as the log-transformed ratio of means (also known as the response ratio), are bounded between 0 and infinity. Finally, some effect sizes, such as the log odds ratio and log risk ratio, require discrete binomial distributions. Let's work through a few examples of how to draw SAFE bootstraps for these effect sizes.

### Multivariate draws

Paired designs offer greater statistical precision and thus have lower standard error and sampling variance. Modifications of common effect sizes account for this. For example, the log-transformed ratio means (here lnRoM), one of the most widely used effect sizes in ecology, is defined as:

$$
lnRoM = log(x_{\text{1}} / x_{\text{2}})
$$ This effect size compares the means of two groups, $x_{\text{1}}$ and {x\_{\text{2}}}. There are delta method derived plug-in formulas for lnRoM that account for paired designs by accepting the variable `r`, or the correlation between the two groups, which is often set at 0.5. These formulas are complicated and difficult to derive for the non-statistician. To use SAFE bootstrapping for a multivariate effect size, one can ignore them and instead draw from a multivariate normal distribution after specifying a covariance matrix. The covariance matrix for a paired lnRoM design is as follows. Note that the multivariate normal distribution function `mvrnorm` in the package `MASS` (or truncated normal, see below) uses variance instead of standard deviation—thus we'll use `sd / n` instead of `sd / sqrt(n)` in the sigma matrix construction.

Also note that the sample sizes of groups 1 and 2 must be equal (unlike the unpaired version of lnRoM).

```{r}
#| label: paired_sigma_matrix
#| echo: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

x1 <- 13.4
x2 <- 16.1
r <- 0.5
sd1 <- 4.6
sd2 <- 3.9
n <- 25

sigma_matrix <- matrix(data = c((sd1^2 / n),        (r*sd1*sd2)/n, #  / n add this to sd1^2
                                (r*sd1*sd2)/n,      (sd2^2 / n)), #  / n add this to sd2^2
                             nrow = 2, ncol = 2)
sigma_matrix

```

One then uses this covariance matrix in their SAFE bootstrap draw:

```{r}
#| label: paired_bootstrap
#| echo: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

cloud <- MASS::mvrnorm(n = 1e6,
                 mu = c(x1, x2),
                 Sigma = sigma_matrix)
cloud <- as.data.frame(cloud)
names(cloud) <- c("x1", "x2")
head(cloud)

```

We can then apply our definition formula to this parameter cloud (step 2 of SAFE) and calculate sampling variance and our bias-corrected point estimates.

```{r}
#| label: paired_SAFE_calculation
#| echo: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"
cloud_trans <- log(cloud$x1 / cloud$x2)

# SAFE sampling variance:
SAFE_vi <- var(cloud_trans)
SAFE_vi

# Plugin estimate from original values:
plugin_effect_size <- log(x1, x2)

SAFE_yi <- (2 * plugin_effect_size) - mean(cloud_trans)

print(c("SAFE bias-corrected point estimate" = SAFE_yi, "SAFE sampling variance" = SAFE_vi))

```

### Truncated normal distribution

Some of the statistics required in effect sizes are bounded, either between 0 and 1, or between 0 and infinity. lnRoM, in our example above, one such effect size. A 0 in the denominator leads Infinity while a zero in the numerator leads to $log(0)$ and -Infinity. To do a SAFE bootstrapping calculation of lnRoM, especially if the mean of a group is near 0, one can nest `rnorm()` inside a `while` loop and filter out any simulated means that are ≤0. Doing this can be computationally expensive though. Alternatively, one can use the much faster `rtmvnorm` function in the `tmvtnorm` package. Simply specify the lower and upper bounds. Note that these must length 2, to match the vector of means (x1 and x2)

```{r}
#| label: truncated_normal
#| echo: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

cloud <- tmvtnorm::rtmvnorm(n = 1e6,
                 mean = c(x1, x2),
                 sigma = sigma_matrix,
                 lower = c(0, 0),
                 upper = c(Inf, Inf))

```

One can then use the same code as above to calculate SAFE point and sampling variance estimates.

### Paired designs
Some effect sizes have paired versions, where the two groups (say control group and treatment group) are paired in some way, because of a before-after design or some other experimental blocking. In these cases $n_1$ and $n_2$ are equal (and so referred to as $n$). Paired design effect sizes are available for lnRoM, SMD, and lnCVR. 

To calculate a paired design effect size, you need to specify an additional parameter: $r$. $r$ is the correlation between group 1 and group 2. Ideally, $r$ would be reported by the study you've digitized. However, this is very uncommon. For that reason, most analysts use several plausible $r$ values (usually 0.5 and 0.8) and then ensure that their overall meta-analytic results are robust to either choice. 

Importantly, the biggest influence of a paired designs is a reduction in sampling variance, which means that paired effect sizes will have more weight in the meta-analytic models. In (@sec-paired-design-monte-carlo) we run SAFE Monte Carlo simulations on several $r$ values and compare their bias to plugin estimates.

To use SAFE for paired designs, you need to do a multivariate draw (e.g., with the `R` functions `MASS::mvrnorm` or `tmvtnorm::rtmvnorm`). This is where things become more complex. Instead of drawing a single point cloud per summary statistic specified with standard error, you draw multiple at once and specify a covariance matrix. 

Consider that the diaganals of this matrix are the standard error of the summary statistics that you are drawing. Let's demonstrate with lnRoM, which is based on a mean for group 1 and group 2 ($lnRoM = log(m_1/m_2)$), with each mean associated with sample size ($n_1$, $n_2$) and standard deviation ($sd_1$, $sd_2$). Because we're interested in drawing a distribution of plausible means, we convert standard deviation to standard error ($sd^2/n$). As such, our **unpaired** matrix looks like this:

$$

Sigma = \begin{bmatrix}
\frac{sd_1^2}{n_1} & 0 \\
0 & \frac{sd_2^2}{n_2}
\end{bmatrix}
$$
The paired version of this would replace the 0s at the top-right and bottom-left corners with terms specifying the covariance between groups 1 and 2. Note that since this is a paired design, $n_1$ and $n_2$ are replaced with $n$


$$
Sigma = \begin{bmatrix}
\frac{sd_1^2}{n} & \frac{r*sd_1*sd_2}{n} \\
\frac{r*sd_1*sd_2}{n} & \frac{sd_2^2}{n}
\end{bmatrix}
$$
Note that if $r$ is 0, then the covariance terms become 0, just as with the unpaired version. 

In `R`, this looks like:

```{r}
m1 <- 10
m2 <- 14.5
sd1 <- 1.3
sd2 <- 1.1
n <- 20
r <- 0.5

sigma_matrix <- matrix(data = c((sd1^2 / n),        (r*sd1*sd2)/n, #  
                                (r*sd1*sd2)/n,      (sd2^2 / n)), #  
                        nrow = 2, ncol = 2)

# Now let's draw our distributions:
out <- MASS::mvrnorm(mu = c(m1, m2),
                    Sigma = sigma_matrix,
                    n = 1e6) |> as.data.frame()
names(out) <- c("m1_cloud", "m2_cloud")
head(out)

```

We can then calculate our bias-corrected point estimates and variance after applying our effect size (lnRoM) to these simulated clouds:

```{r}

cloud_trans <- log(out$m1_cloud / out$m2_cloud)

vi_SAFE <- var(cloud_trans)
  
plugin_effect_size <- log(m1 / m2)

# Bias corrected SAFE estimate:
yi_SAFE <- (2 * plugin_effect_size) - mean(cloud_trans)

c("yi_SAFE" = yi_SAFE, "vi_SAFE" = vi_SAFE)

```

See below (@sec-paired-design-monte-carlo) for a simulation on how different values of $r$ affect the bias of these estimates.

### Binomial distributions

Many effect sizes have discrete outcome variables, e.g., number of patients with some outcome. The log odds ratio (lnOR) is a commonly used effect size for these types of data. The formula is:

$$
lnOR = log((a * d) / (b * c))
$$
Where groups a, b, c, d each reflect a different combination of treatments and outcomes. For example, groups 'a' and 'b' are grasshoppers exposed to a pesticide, while groups 'c' and 'd' were given a placebo. 'b' and 'd' are the number of grasshoppers that died, while groups 'a' and 'c' are the number of grasshoppers who lived.

|           | living | dead | total          |
|-----------|--------|------|----------------|
| pesticide | $a$    | $b$  | $n_{\text{1}}$ |
| control   | $c$    | $d$  | $n_{\text{2}}$ |

To conduct SAFE on a binomial distribution like these, one uses 'rbinom()' to draw 'a' and 'c' based on the probability of 'a' and 'c' (e.g., $a / n_{\text{1}}$ and $c / n_{\text{2}}$).

```{r}
#| label: binomial_draw
#| echo: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"
a <- 3
c <- 8
b <- 7
d <- 2

# Row totals:
n1 <- a + b
n2 <- c + d

# Probability of a and c:
pr_a <- a/n1
pr_c <- c/n2

a_cloud <- rbinom(1e6, n1, pr_a)
c_cloud <- rbinom(1e6, n2, pr_c)

b_cloud <- n1 - a_cloud
d_cloud <- n2 - c_cloud

cloud_trans = log((a_cloud * d_cloud) / (b_cloud * c_cloud))

```

One can then estimate sampling variance and bias-corrected point estimates as with all other effect sizes:

```{r}
#| label: lnOR_SAFE
#| echo: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

vi_SAFE <- var(cloud_trans)

plugin_effect_size <- log((a * d) / (b * c))

SAFE_yi <- (2 * plugin_effect_size) - mean(cloud_trans)

print(c("SAFE bias-corrected point estimate" = SAFE_yi, "SAFE sampling variance" = SAFE_vi))

```

# How good is SAFE? {#sec-compare-results}

We simulated the bias of SAFE calculations versus normal plug-in calculations for a variety of common, and less common, effect sizes. We did this across a range of input variables, particularly sample size and the number of SAFE bootstraps.

In these simulations, we created simulated datasets based on true values for each scenario. These true values are the *population means*, of which each *sample mean* is drawn from based on the scenario sample size. We then calculated effect sizes and sampling variances (both with plug-in formulas and SAFE) for the true values and based on the simulated datasets. We did this 1e5 times for each scenario.

In these Monte Carlo simulations, we were interested in bias, or the difference between the 'true' estimands (i.e., sampling variance and effect sizes) and the estimates (of sampling variance and effect sizes) produced from various methods and applied to the simulated data. We'll explain the difference between estimands, estimators, and estimates below.

To illustrate, we'll do a simulation for our 'speed' effect size `1/x`. Note that we'll do this for only 1 effect size (e.g., mean, SD combination) but across several sample sizes, to illustrate how bias decreases with sample size.

To speed things up we will do this in parallel. The function `prepare_cluster()` below can help set things up and includes a progress bar so you don't lose your mind wondering what's going on in there.

```{r, prepare_cluster}
#| label: prepare_cluster
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

set.seed(2025)

# Prepare cluster:
prepare_cluster <- function(n){
    require("parallel")
    require("foreach")
    require("doSNOW")
    
    nCores <- parallel::detectCores() -1 
    cl <- makeCluster(nCores)
    registerDoSNOW(cl)
    
    # Progress bar
    pb <- txtProgressBar(max = n, style = 3)
    progress <- function(n) setTxtProgressBar(pb, n)
    opts <- list(progress = progress)
    ret <- list(opts, pb, cl)
    names(ret) <- c("options", "progress", "cluster")
    
    return(ret)
    
    cat("Pass 'x$options' to .opts in foreach;
      'x$progress' to setTxtProgressBar(x$progress, i);
      'x$cluster' to stopCluster(x$cluster) after foreach")
}

```

Now we will run 100,000 simulations to demonstrate how this method works. Simulation results for the other effect sizes are reported below (@sec-Monte-Carlo-results).

To begin, let's imagine that a population has a true *population mean* latency time of 15 and a true standard deviation of 1.2. This population was sampled with varying degrees of effort, and thus has sample sizes of 5, 15, and 100. These sample sizes will be used to create our simulated datasets.

```{r}
#| label: scenario
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

guide <- data.table(true_mean = 10.5,
                    true_sd = 1.2,
                    sample_n = c(5, 15, 100),
                    scenario_id = paste0("scenario_", seq(1:3)))
guide

```

### Monte Carlo simulation

We'll run this Monte Carlo simulation in a foreach loop for 1e5 iterations. Note that since we're doing this simulation for 3 different sample sizes, we need to do the simulation in an `lapply` inside our foreach loop. In this lapply, we'll create simulated datasets sampled 5, 15, and 100 based on the 'true' mean and standard deviation.

This simulated data is then summarized to a *sample mean* and standard deviation from which we will calculate effect sizes and sampling variances with plug-in and SAFE methods. This is approximating reality, wherein researchers are estimating a 'true' value in the world by taking samples of it, and the job of the meta-analyst is to compare their results while accounting for differences in sampling.

Note that if you're running this code interactively, this can take a minute. Feel free to load the already run dummy dataset in the next code block.

```{r, monte_carlo}
#| label: monte_carlo
#| eval: false
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"
  
monte_carlo_N <- 1e5

clust_out <- prepare_cluster(n = monte_carlo_N)

res <- foreach(i = 1:monte_carlo_N, 
       .options.snow = clust_out$options,
       .errorhandling = "pass",
       .packages = c("data.table", "MASS", "tmvtnorm")) %dopar% {
         
  # This lapply will cycle through each scenario:
 results  <- lapply(1:nrow(guide), function(x){
   
   # Draw sample data with standard deviation and sample_n:
   sample <- rnorm(guide$sample_n[x],               # number of samples
                     mean = guide$true_mean[x],     # true mean latency time
                     sd = guide$true_sd[x])         # standard deviation

   # Calculate sample mean/sd
   sim_dat <- data.table(sim_mean = mean(sample),
                         sim_sd = sd(sample),
                         sample_n = length(sample),
                         true_mean = guide$true_mean[x],
                         true_sd = guide$true_sd[x],
                         scenario_id = guide$scenario_id[x])
   
    # Now calculate plug-in effect sizes and sampling variances for each simulation sample:
   sim_dat[, yi_plugin := 1/sim_mean]
   sim_dat[, vi_plugin := (sim_sd^2 / sample_n) * (-1 * sim_mean ^ -2)^2]

   # Now, calculate SAFE effect sizes and sampling variances. 
   # First, create cloud of hyperparameters:
   cloud <- rnorm(1e5,
                  mean = sim_dat$sim_mean,    
                  sd = sim_dat$sim_sd / sqrt(sim_dat$sample_n))
   
   # Transform cloud:
   cloud_trans <- 1/cloud
   
   # Calculate SAFE sample variance and add to our 'sim_dat'
   sim_dat[, vi_safe := var(cloud_trans)]
   
   # Calculate bias corrected SAFE point estimate and add to our 'sim_dat'
   sim_dat[, yi_safe := (2 * yi_plugin) - mean(cloud_trans)]
   
   # And, for demonstration purposes, let's also record the uncorrected mean of our 'cloud_trans'
   sim_dat[, mean_safe := mean(cloud_trans)]

   # Return the results:
   return(sim_dat)
 }) |> rbindlist()
 
 setTxtProgressBar(clust_out$progress, i)
 
 return(results)

}
stopCluster(clust_out$cluster)

res <- rbindlist(res)  

saveRDS(res, "builds/speed_simulation.Rds")

```

Or just load the simulation here:

```{r, monte_carlo_load}
#| label: monte_carlo_load
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

res <- readRDS("builds/speed_simulation.Rds")

head(res, 3)

```

Now, let's add the 'true' effect size, based on the underlying *population mean* and *SD* to this dataset:

```{r, true_values}
#| label: true_values
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

res[, true_yi := 1/true_mean]

```

### Calculate bias

To interpret the simulation results, it is essential to understand the difference between *estimands*, *estimators*, and *estimates*.

**Estimands** are the 'true' point estimates and sampling variances.

**Estimators** are the methods used to estimate the estimand. What a tongue twister! In our case, the estimators are the 1st order effect size point estimate (e.g., definition formula), the sampling variance formula derived via the delta method (`(sd^2 / n) * (-1 * x ^ -2)^2`), and the SAFE method point estimates and sampling variances..

**Estimates** are the estimates of the estimands produced by the estimators 🤠

### Effect size (point estimate) bias

In our example, there are two estimands that we're interested in estimating: the 'true' effect size (based on the definition formula of the effect size type) and the 'true' sampling variance. To calculate the bias of our various estimators in estimating effect sizes, we calculate the mean of the estimates from each simulated dataset and subtract the estimand value (the true effect size off the 'true' values).

Here we will calculate the bias of 3 point estimate estimators: the plug-in (`yi_plugin`), the bias-corrected SAFE estimate (`yi_safe`), and the uncorrected SAFE mean (`mean_safe`):

```{r, calc_bias}
#| label: calc_bias
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

bias <- function(estimates, estimand){
  return(mean(estimates) - unique(estimand))
}

point.bias <- res[, .(plugin_bias = bias(yi_plugin, true_yi),
                      bias_corrected_safe_bias = bias(yi_safe, true_yi),
                      uncorrected_safe_bias = bias(mean_safe, true_yi)),
                  by = .(scenario_id, sample_n)] |> unique()
head(point.bias)

# Melt this to make it plottable:
point.bias.long <- melt(point.bias,
                        id.vars = c("scenario_id", "sample_n"),
                        value.name = "bias",
                        variable.name = "estimator")

head(point.bias.long)

# Sort by sample size 
setorder(point.bias.long, sample_n)

ggplot(data = point.bias.long,
       aes(x = sample_n, y = bias, color = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(linewidth = 1)+
  geom_point(size = 2)+
  xlab("Sample size")+
  ylab("Point estimate bias")+
  coord_cartesian(ylim = c(-0.001, 0.001))+
  scale_color_manual("Estimator",
                       labels = c("plugin_bias" = "Plug-in bias",
                                  "bias_corrected_safe_bias" = "SAFE\n(bias-corrected)",
                                  "uncorrected_safe_bias" = "mean SAFE\n(uncorrected)"),
                       values = c("plugin_bias" = "#A8DADC",
                                   "bias_corrected_safe_bias" = "#E63946",
                                   "uncorrected_safe_bias" = "#1D3557"))+
  theme_bw()+
  theme(panel.border = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom")

```

Wow! It looks like the bias-corrected SAFE estimate is the least biased estimator of our true effect size.

### Sampling variance bias

Now, let's calculate the bias of our sampling variance estimates. Interestingly, there is no way to know the 'true' estimand for sampling variance because sampling variance itself is a product of sampling (e.g., there is no true value). To estimate bias for our sampling variance estimates, we actually calculate the variance in point estimates for each estimator and use *each* of these as the estimand. With 2 estimators and thus 2 estimands, we'll end up with 4 calculations of bias.

To understand how well our approximation is performing for sampling variance, we typically calculate *relative bias* by calculating bias as above but then dividing by the estimand and multiplying by 100.

```{r, calc_rel_bias}
#| label: calc_rel_bias
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

# First, calculate variance in point estimates as our 'estimand' by scenario_id
res[, var_estimand_plugin := var(yi_plugin), 
    by = .(scenario_id, sample_n)]
res[, var_estimand_SAFE := var(yi_safe), 
    by = .(scenario_id, sample_n)]

# To make this easier to read, we'll encapsulate the relative bias in a function:
relative_bias <- function(estimates,
                          estimand){
  ((mean(estimates) - estimand ) / estimand) * 100
}

# Now summarize by calculating relative bias per scenario ID and sample size
var.bias <- res[, .(SAFE_estimate.plugin_estimand = relative_bias(vi_safe, unique(var_estimand_plugin)),
                    SAFE_estimate.SAFE_estimand = relative_bias(vi_safe, unique(var_estimand_SAFE)),
                    plugin_estimate.plugin_estimand = relative_bias(vi_plugin, unique(var_estimand_plugin)),
                    plugin_estimate.SAFE_estimand = relative_bias(vi_plugin, unique(var_estimand_SAFE))),
    by = .(scenario_id, sample_n)] |> unique()

# Now melt:
var.bias.long <- melt(var.bias,
                      id.vars = c("scenario_id", "sample_n"))
head(var.bias.long)

# Split the 'variable' into estimator and estimand for plotting
var.bias.long[, c("Estimator", "Estimand") := tstrsplit(variable, ".", fixed = TRUE)]

# Sort the dataset by sample size
setorder(var.bias.long, sample_n)

# Plot
ggplot(data = var.bias.long, 
       aes(x = sample_n, y = value, color = Estimator))+
  geom_hline(yintercept = 0)+
  geom_path(linewidth = 1)+
  geom_point(size = 2)+
  xlab("Sample size")+
  ylab("Relative bias (%) in variance estimates")+
  scale_color_manual("Estimator",
                     labels = c("plugin_estimate" = "Plug-in\n(definition formula)",
                                "SAFE_estimate" = "SAFE bootstrapping"),
                     values = c("plugin_estimate" = "#A8DADC",
                                "SAFE_estimate" = "#E63946"))+
  facet_wrap(~Estimand,
             ncol = 1,
             labeller = as_labeller(c("plugin_estimand" = "Plug-in estimand\n(definition formula)",
                                      "SAFE_estimand" = "SAFE estimand")))+
  theme_bw()+
  theme(panel.border = element_blank(),
        panel.grid = element_blank(),
        strip.background = element_blank(),
        legend.position = "bottom")

```

We see here that SAFE (red) is slightly more biased than the plug-in variance formula we derived. However, the difference is marginal and more conservative, given that higher sampling variance means lower weights in the meta-analytic model.

We've conducted full Monte Carlo simulations for a variety of effect sizes, which are below (@sec-Monte-Carlo-results). However, first, we'll demonstrate how a meta-analyst might use SAFE to calculate point estimates and sampling variances for a custom effect size.

## Full Monte Carlo results {#sec-Monte-Carlo-results}

Now let's look at the simulation results for a variety of effect sizes.

```{r, sim_all}
#| label: sim_all
#| echo: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

sim_results <- readRDS("builds/all_scenarios_summarized.Rds")
sim_results <- sim_results[boots == 1e6, ]

```

### Speed

These are the full results of our 'speed' (`1/x`) effect size Monte Carlo simulation.

```{r, themes}
#| label: themes
#| echo: false
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

var_lab <- "Relative bias of\nvariance estimates (%)"
point_lab <- "Bias of point\nestimates"

theme_SAFE <- theme_bw()+
  theme(panel.border = element_blank(),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        panel.grid = element_blank(),
        strip.background = element_blank())

labs = c("plugin_1st" = "1st order plugin",
           "plugin_2nd" = "2nd order plugin",
           "safe" = "SAFE")

pal = c("plugin_1st" = "#A8DADC",
           "plugin_2nd" = "#1D3557",
           "safe" = "#E63946")
pt_size <- 2

facet_labs <- as_labeller(c("plugin_1st" = "1st order estimand",
                            "plugin_2nd" = "2nd order plugin estimand",
                            "safe" = "SAFE estimand"))
```

```{r, rec_speed}
#| label: rec_speed
#| echo: false
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"
sub_dat <- sim_results[effect_type == "reciprocal", ]

setorder(sub_dat, sample_size)

p.point <- ggplot(data = sub_dat[calculation %in% c("bias"), ], 
            aes(x = sample_size, y = value, color = estimator,
                group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(point_lab)+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs)+
  facet_wrap(~estimand, 
             labeller = facet_labs)+
  coord_cartesian(ylim = c(-0.0003, 0.0003))+
  ggtitle("Speed (1 / x)")+
  theme_SAFE+
  theme(legend.position = "none") 

p.variance <- ggplot(data = sub_dat[calculation %in% c("relative_bias"), ], 
             aes(x = sample_size, y = value, color = estimator,
                 group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(var_lab)+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs)+
  facet_wrap(~estimand, 
             labeller = facet_labs,
             nrow = 2)+
  coord_cartesian(ylim = c(-5, 5))+
  theme_SAFE+
  theme(legend.position = "bottom")

p.point + p.variance + plot_layout(ncol = 1, heights = c(1/3, 2/3)) + 
  plot_annotation(tag_levels = "A")

```

It looks like the SAFE point estimate is *less* biased than the plugin estimate! It also looks like the SAFE variance estimate is more conservative than the plugin variance estimate, which means that meta-analytic data won't have more weight than it should in a meta-analytic model.

An additional dimension to check is **coverage**, which describes whether the SAFE estimate ± CIs includes the 'true' point estimand. We'll plot the proportion of Monte Carlo simulations where the true point estimand was inside the SAFE and plugin estimates ± CIs:

```{r}
#| label: coverage_speed
#| echo: false
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

p.coverage <- ggplot(data = sub_dat[calculation %in% c("coverage"), ], 
            aes(x = sample_size, y = value, color = estimator,
                group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab("Coverage")+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs)+
  facet_wrap(~estimand, 
             labeller = facet_labs)+
  coord_cartesian(ylim = c(.93, 1))+
  ggtitle("Speed (1 / x)")+
  theme_SAFE+
  theme(legend.position = "bottom") 
p.coverage

```

Coverage appears virtually identically for SAFE and the plugin at moderate sample sizes, but note that SAFE has better coverage at low sample sizes. That's pretty cool.


### lnRoM

Now let's look at the log-transformed ratio of means (lnRoM or lnRR). 

```{r, SMD_mvnormal}
#| label: lnRoM
#| echo: false
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

sub_dat <- sim_results[effect_type == "lnRoM", ]

setorder(sub_dat, sample_size1)

p.point <- ggplot(data = sub_dat[calculation %in% c("bias") & sample_size_ratio == 1,],
                  aes(x = sample_size1, y = value, color = estimator,
                      group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(point_lab)+
  xlab("Sample size")+
  ggtitle("SMD", subtitle= "4-multivariate normal")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs)+
  facet_wrap(~estimand, 
             labeller = as_labeller(c("plugin_1st" = "1st order plugin estimand",
                                      "plugin_2nd" = "2nd order plugin estimand")))+
  coord_cartesian(ylim = c(-0.01, 0.01))+
  ggtitle("SMD")+
  theme_SAFE+
  theme(legend.position = "none") 

#
p.variance <- ggplot(data = sub_dat[calculation %in% c("relative_bias") & sample_size_ratio == 1, ], 
                     aes(x = sample_size1, y = value, color = estimator,
                         group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(var_lab)+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = c(plugin_1st = "1st order plugin", 
                                plugin_2nd = "2nd order plugin", 
                                safe = "SAFE"))+
  coord_cartesian(ylim = c(-30, 30))+
  facet_wrap(~estimand, 
             ncol = 1,
             labeller = as_labeller(c("plugin_1st" = "1st order plugin estimand",
                                      "plugin_2nd" = "2nd order plugin estimand",
                                      "safe" = "SAFE estimand")))+
  theme_SAFE+
  theme(legend.position = "bottom")

p.point + xlab(NULL) + p.variance + plot_layout(ncol = 1, heights = c(1/4, 3/4)) + 
  plot_annotation(tag_levels = "A")

```

And coverage:
```{r}
#| label: coverage_smd_normal
#| echo: false
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

p.coverage <- ggplot(data = sub_dat[calculation %in% c("coverage"), ], 
            aes(x = sample_size1, y = value, color = estimator,
                group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab("Coverage")+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs)+
  facet_wrap(~estimand, 
             labeller = facet_labs,
             ncol = 2)+
  coord_cartesian(ylim = c(.95, .99))+
  theme_SAFE+
  theme(legend.position = "bottom") 
p.coverage

```
Slightly better coverage than the plugin estimates! Neat 🙂

### SMD 4 multivariate normal

The SAFE estimator for this effect size was calculated with 4 multivariate normal distributions for both mean and SD, as both of these parameters are in the definition formula (Cohen's d) of SMD. Let's compare the bias of SAFE, Cohen's d, and Hedges' g.

```{r, SMD_mvnormal}
#| label: SMD_mvnormal
#| echo: false
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

sub_dat <- sim_results[effect_type == "SMD_normal", ]

setorder(sub_dat, sample_size1)

p.point <- ggplot(data = sub_dat[calculation %in% c("bias") & sample_size_ratio == 1,],
                  aes(x = sample_size1, y = value, color = estimator,
                      group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(point_lab)+
  xlab("Sample size")+
  ggtitle("SMD", subtitle= "4-multivariate normal")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs)+
  facet_wrap(~estimand, 
             labeller = as_labeller(c("plugin_1st" = "Cohen's d estimand")))+
  coord_cartesian(ylim = c(-0.1, 0.1))+
  ggtitle("SMD")+
  theme_SAFE+
  theme(legend.position = "none") 

#
p.variance <- ggplot(data = sub_dat[calculation %in% c("relative_bias") & sample_size_ratio == 1, ], 
                     aes(x = sample_size1, y = value, color = estimator,
                         group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(var_lab)+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = c(plugin_1st = "Cohen's d", 
                                plugin_2nd = "Hedges' g", 
                                safe = "SAFE"))+
  coord_cartesian(ylim = c(-30, 30))+
  facet_wrap(~estimand, 
             ncol = 1,
             labeller = as_labeller(c("plugin_1st" = "Cohen's d estimand",
                                      "plugin_2nd" = "Hedges' g estimand",
                                      "safe" = "SAFE estimand")))+
  theme_SAFE+
  theme(legend.position = "bottom")

p.point + xlab(NULL) + p.variance + plot_layout(ncol = 1, heights = c(1/4, 3/4)) + 
  plot_annotation(tag_levels = "A")

```

And coverage:
```{r}
#| label: coverage_smd_normal
#| echo: false
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

p.coverage <- ggplot(data = sub_dat[calculation %in% c("coverage"), ], 
            aes(x = sample_size1, y = value, color = estimator,
                group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab("Coverage")+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs)+
  facet_wrap(~estimand, 
             labeller = facet_labs,
             ncol = 2)+
  coord_cartesian(ylim = c(.95, 1))+
  theme_SAFE+
  theme(legend.position = "bottom") 
p.coverage

```


### SMD 2 multivariate normal and 2-Wishart

With this estimator, the SAFE calculation used the normal distribution to simulate the two means but the Wishart distribution to estimate the SD.

```{r, SMD_Wishart}
#| label: SMD_Wishart
#| echo: false
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

sub_dat <- sim_results[effect_type == "SMD_Wishart", ]
setorder(sub_dat, sample_size1)

p.point <- ggplot(data = sub_dat[calculation %in% c("bias") & sample_size_ratio == 1, ],
                  aes(x = sample_size1, y = value, color = estimator,
                      group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(point_lab)+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator",
                     values = pal,
                     labels = labs)+
  facet_wrap(~estimand,
             labeller = as_labeller(c("plugin_1st" = "Cohen's d estimand")))+
  coord_cartesian(ylim = c(-0.1, 0.1))+
  ggtitle("SMD", subtitle= "2-multivariate normal, 2-multivariate Wishart")+
  theme_SAFE+
  theme(legend.position = "none")

#
p.variance <- ggplot(data = sub_dat[calculation %in% c("relative_bias") & sample_size_ratio == 1, ],
                     aes(x = sample_size1, y = value, color = estimator,
                         group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(var_lab)+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator",
                     values = pal,
                     labels = c(plugin_1st = "Cohen's d",
                                plugin_2nd = "Hedges' g",
                                safe = "SAFE"))+
  facet_wrap(~estimand,
             ncol = 1,
             labeller = as_labeller(c("plugin_1st" = "Cohen's d estimand",
                                      "plugin_2nd" = "Hedges' g estimand",
                                      "safe" = "SAFE estimand")))+
  coord_cartesian(ylim = c(-60, 60))+
  theme_SAFE+
  theme(legend.position = "bottom")

p.point + xlab(NULL) + p.variance + plot_layout(ncol = 1, heights = c(1/4, 3/4)) + 
  plot_annotation(tag_levels = "A")

```


And coverage:
```{r}
#| label: coverage_smd_wishart
#| echo: false
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

p.coverage <- ggplot(data = sub_dat[calculation %in% c("coverage"), ], 
            aes(x = sample_size1, y = value, color = estimator,
                group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab("Coverage")+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs)+
  facet_wrap(~estimand, 
             labeller = facet_labs,
             ncol = 2)+
  coord_cartesian(ylim = c(.95, 1))+
  theme_SAFE+
  theme(legend.position = "bottom") 
p.coverage

```
😎

### lnCVR 4 multivariate normal

The SAFE estimator for this effect size was calculated with 4 multivariate normal distributions for both mean and SD.

```{r, lnCVR_mvnormal}
#| label: lnCVR_mvnormal
#| echo: false
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

sub_dat <- sim_results[effect_type == "lnCVR_normal", ]
setorder(sub_dat, sample_size1)

p.point <- ggplot(data = sub_dat[calculation %in% c("bias") & 
                                   sample_size_ratio == 1, ], 
                  aes(x = sample_size1, y = value, color = estimator,
                      group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(point_lab)+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs)+
  facet_wrap(~estimand, 
             labeller = facet_labs)+
  coord_cartesian(ylim = c(-0.02, 0.02))+
  ggtitle("lnCVR", subtitle = "4-multivariate normal")+
  theme_SAFE+
  theme(legend.position = "none") 

#
p.variance <- ggplot(data = sub_dat[calculation %in% c("relative_bias") & sample_size_ratio == 1, ], 
                     aes(x = sample_size1, y = value, color = estimator,
                         group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(var_lab)+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs)+
  facet_wrap(~estimand, 
             ncol = 1,
             labeller = facet_labs)+
  coord_cartesian(ylim = c(-60, 60))+
  theme_SAFE+
  theme(legend.position = "bottom")

p.point + xlab(NULL) + p.variance + plot_layout(ncol = 1, heights = c(1/4, 3/4)) + 
  plot_annotation(tag_levels = "A")

```

And coverage:
```{r}
#| label: coverage_lncvr_normal
#| echo: false
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

p.coverage <- ggplot(data = sub_dat[calculation %in% c("coverage"), ], 
            aes(x = sample_size1, y = value, color = estimator,
                group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab("Coverage")+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs)+
  facet_wrap(~estimand, 
             labeller = facet_labs,
             ncol = 2)+
  coord_cartesian(ylim = c(.95, 1))+
  theme_SAFE+
  theme(legend.position = "bottom") 
p.coverage

```

### lnCVR 2 multivariate normal and 2-Wishart

The SAFE estimator for this effect size was calculated with 2 multivariate normal distributions for the means and 2 Wishart distributions for the SD.

```{r, lnCVR_Wishart}
#| label: lnCVR_Wishart
#| echo: false
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

sub_dat <- sim_results[effect_type == "lnCVR_Wishart", ]
setorder(sub_dat, sample_size1)

p.point <- ggplot(data = sub_dat[calculation %in% c("bias") &
                                   sample_size_ratio == 1, ],
                  aes(x = sample_size1, y = value, color = estimator,
                      group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(point_lab)+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator",
                     values = pal,
                     labels = labs)+
  coord_cartesian(ylim = c(-0.02, 0.02))+
  facet_wrap(~estimand,
             labeller = facet_labs)+
    ggtitle("lnCVR", subtitle = "2-multivariate normal, 2-multivariate Wishart")+
  theme_SAFE+
  theme(legend.position = "none") 

#
p.variance <- ggplot(data = sub_dat[calculation %in% c("relative_bias") & sample_size_ratio == 1, ],
                     aes(x = sample_size1, y = value, color = estimator,
                         group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(var_lab)+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator",
                     values = pal,
                     labels = labs)+
  coord_cartesian(ylim = c(-20, 20))+
  facet_wrap(~estimand,
             ncol = 1,
             labeller = facet_labs)+
  theme_SAFE+
  theme(legend.position = "bottom")

p.point + xlab(NULL) + p.variance + plot_layout(ncol = 1, heights = c(1/4, 3/4)) + 
  plot_annotation(tag_levels = "A")

```

And coverage:
```{r}
#| label: coverage_lnCVR_wishart
#| echo: false
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

p.coverage <- ggplot(data = sub_dat[calculation %in% c("coverage"), ], 
            aes(x = sample_size1, y = value, color = estimator,
                group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab("Coverage")+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs)+
  facet_wrap(~estimand, 
             labeller = facet_labs,
             ncol = 2)+
  coord_cartesian(ylim = c(.95, 1))+
  theme_SAFE+
  theme(legend.position = "bottom") 
p.coverage

```


### Log Odds Ratio (lnOR)

In this effect size estimation, the SAFE calculation used a binomial distribution to draw random binomial samples for 'a' and 'c'.

```{r, lnOR}
#| label: lnOR
#| echo: false
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

sub_dat <- sim_results[effect_type == "lnOR", ]
setorder(sub_dat, n1)

p.point <- ggplot(data = sub_dat[calculation %in% c("bias"), ], 
                  aes(x = n1, y = value, color = estimator,
                      group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(point_lab)+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs)+
  facet_wrap(~estimand,
             labeller = facet_labs)+
  coord_cartesian(ylim = c(-.25, .25))+
  ggtitle("lnOR")+
  theme_SAFE+
  theme(legend.position = "none") 

#
p.variance <- ggplot(data = sub_dat[calculation %in% c("relative_bias"), ], 
                     aes(x = n1, y = value, color = estimator,
                         group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(var_lab)+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs)+
  facet_wrap(~estimand, 
             ncol = 1,
             labeller = facet_labs)+
  coord_cartesian(ylim = c(-30, 30))+
  theme_SAFE+
  theme(legend.position = "bottom")

p.point + xlab(NULL) + p.variance + plot_layout(ncol = 1, heights = c(1/4, 3/4)) + 
  plot_annotation(tag_levels = "A")

```

And coverage:
```{r}
#| label: coverage_lnOR
#| echo: false
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

p.coverage <- ggplot(data = sub_dat[calculation %in% c("coverage"), ], 
            aes(x = n1, y = value, color = estimator,
                group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab("Coverage")+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs)+
  facet_wrap(~estimand, 
             labeller = facet_labs,
             ncol = 2)+
  coord_cartesian(ylim = c(.95, 1))+
  theme_SAFE+
  theme(legend.position = "bottom") 
p.coverage

```


### Log Risk Ratio (lnRR)

In this effect size estimation, the SAFE calculation drew 'a' and 'c' from a binomial distribution.

```{r, lnRR}
#| label: lnRR
#| echo: false
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

sub_dat <- sim_results[effect_type == "lnRR", ]
setorder(sub_dat, n1)

p.point <- ggplot(data = sub_dat[calculation %in% c("bias") , ], 
                  aes(x = n1, y = value, color = estimator,
                      group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(point_lab)+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs)+
  facet_wrap(~estimand, 
             labeller = facet_labs)+
  coord_cartesian(ylim = c(-.2, .2))+
  ggtitle("lnRR")+
  theme_SAFE+
  theme(legend.position = "none") 

#
p.variance <- ggplot(data = sub_dat[calculation %in% c("relative_bias"), ], 
                     aes(x = n1, y = value, color = estimator,
                         group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(var_lab)+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs)+
  facet_wrap(~estimand, 
             ncol = 1,
             labeller = facet_labs)+
  coord_cartesian(ylim = c(-40, 40))+
  theme_SAFE+
  theme(legend.position = "bottom")

p.point + xlab(NULL) + p.variance + plot_layout(ncol = 1, heights = c(1/4, 3/4)) + 
  plot_annotation(tag_levels = "A")

```

And coverage:
```{r}
#| label: coverage_lnRR
#| echo: false
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

p.coverage <- ggplot(data = sub_dat[calculation %in% c("coverage"), ], 
            aes(x = n1, y = value, color = estimator,
                group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab("Coverage")+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs)+
  facet_wrap(~estimand, 
             labeller = facet_labs,
             ncol = 2)+
  coord_cartesian(ylim = c(.95, 1))+
  theme_SAFE+
  theme(legend.position = "bottom") 
p.coverage

```
Wowza!


### Hardy Weinberg Disequilibrium

In this effect size estimation, the SAFE calculation drew 'n_AA', 'n_Aa', and 'n_aa' from a binomial distribution.

```{r, lnHWE_A}
#| label: lnHWE_A
#| echo: false
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

sub_dat <- sim_results[effect_type == "lnHWE_A", ]
setorder(sub_dat, n)

p.point <- ggplot(data = sub_dat[calculation %in% c("bias"), ], 
                  aes(x = n, y = value, color = estimator,
                      group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(point_lab)+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs)+
  facet_wrap(~estimand, 
             labeller = facet_labs)+
  coord_cartesian(ylim = c(-0.1, 0.1))+
  ggtitle("Hardy-Weinburg Disequilibrium")+
  theme_SAFE+
  theme(legend.position = "none") 

#
p.variance <- ggplot(data = sub_dat[calculation %in% c("relative_bias") & 
                                   p_AA == 0.25 & p_aa == 0.25 & p_Aa == 0.5, ], 
                     aes(x = n, y = value, color = estimator,
                         group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(var_lab)+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs)+
  facet_wrap(~estimand, 
             ncol = 1,
             labeller = facet_labs)+
    coord_cartesian(ylim = c(-15, 15))+
  theme_SAFE+
  theme(legend.position = "bottom")


p.point + xlab(NULL) + p.variance + plot_layout(ncol = 1, heights = c(1/4, 3/4)) + 
  plot_annotation(tag_levels = "A")

```


And coverage:
```{r}
#| label: coverage_hwe
#| echo: false
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

p.coverage <- ggplot(data = sub_dat[calculation %in% c("coverage"), ], 
            aes(x = n, y = value, color = estimator,
                group = estimator))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab("Coverage")+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs)+
  facet_wrap(~estimand, 
             labeller = facet_labs,
             ncol = 2)+
  coord_cartesian(ylim = c(.92, 1))+
  theme_SAFE+
  theme(legend.position = "bottom") 
p.coverage
```


# How many SAFE bootstraps?

The SAFE method relies on bootstrapping to calculate effect size point estimates and sampling variance. It's sort of magical how well it works! But, how many bootstraps are necessary? Let's find out by doing another Monte Carlo simulation, this time of lnRoM (log-transformed ratio of means), which is one of the most widely used effect sizes in ecology. This time, instead of calculating bias, we'll look at the standard deviation of the point and sampling variance estimates as we change the number of bootstraps.

However, before we go any further, we'll load a function (`eff_size`) we wrote to calculate effect sizes both with plug-in estimators and with SAFE. This function can be sourced into the local environment. Let's load the function and then calculate effect sizes, with SAFE, for several examples from the main text. The function is located in the github repo at scripts/SAFE_function.R.

### Load functions

This function returns a data.table (basically a *fast* data.frame) with columns for plug-in effect sizes and sampling variances (denoted with `_first` or `_second` based on derivative order) and SAFE effect sizes and sampling variances (denoted with `_safe`). Point estimates (effect sizes) are denoted with `yi_` while sampling variances are denoted with `vi_`.

Here, we'll demonstrate with the effect size 'lnRoM', or log-transformed ratio of means, which is a commonly used effect size to compare control and experimental groups.

```{r, safe_function_loRom}
#| label: safe_function_loRom
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

source("scripts/SAFE_function.R")

# Set seed for reproducibility
set.seed(2025)

# For a single effect size and sampling variance:
eff_size(x1 = 14.5, x2 = 7.9, 
         sd1 = 1.3, sd2 = 2,
         n1 = 20, n2 = 20, 
         effect_type = "lnRoM",
         SAFE = TRUE,
         verbose = TRUE,
         SAFE_boots = 1e6)

```

The function also works with a vector of raw data:

```{r, safe_function_loRom_raw}
#| label: safe_function_loRom_raw
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

eff_size(x1 = c(14.5, 13, 15.8), 
         x2 = c(7.9, 21, 18.4), 
         sd1 = c(1.3, 2, 1.9), 
         sd2 = c(2, 3.1, 1.4),
         n1 = c(20, 20, 20), 
         n2 = c(20, 15, 18), 
         effect_type = "SMD",
         SAFE = TRUE,
         verbose = FALSE,
         parallelize = TRUE,
         SAFE_boots = 1e6)

```

For a full list of options that `eff_size` can calculate, run:

```{r, eff_sizes_lookup}
#| label: eff_sizes_lookup
#| eval: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

eff_size()

```

To store results simply assign the output to an object in R. You will then be able to access the results as a data.table.

::: callout-tip
### eff_size() function arguments

-   `...`: Named vectors (numeric) of the variables required by the selected effect size (run `eff_size()` for a list of accepted effect sizes). All supplied vectors must have identical lengths. The required variable names depend on `effect_type` and must match the `vars_required` field for that effect in `data/effect_size_formulas.csv` (see `eff_size()`). If a paired design is requested and `r` is not supplied, it defaults to 0.5; for non-paired designs, `r` defaults to 0.

-   `effect_type`: Character (scalar). Name of the effect size to compute, matching a `name` value in `data/effect_size_formulas.csv` (run `eff_size()` for accepted effect size names). If no effect size name is provided, the function returns a lookup table of available effect sizes and their required variables and exits.

-   `SAFE`: Logical. If `TRUE` (default), perform SAFE bias correction using parametric bootstrap draws specified by `SAFE_boots`. If `FALSE`, only the plug-in effect(s) are returned.

-   `SAFE_boots`: Integer. Number of bootstrap samples used by SAFE (10\^6 by default). Larger values improve precision but increase run time. Used only when `SAFE = TRUE`.

-   `parallelize`: Logical. If `TRUE` (default), SAFE calculations are parallelized over observations via `parallel::mclapply`, using `detectCores() - 1` workers. Set to `FALSE` to run SAFE serially. Note: `mclapply` uses forking (Unix/macOS); on Windows this may fall back to single-core behavior.

-   `verbose`: Logical. If `TRUE` (default), prints the selected formulas, special warnings attached to the effect type (e.g., handling of zeros or negatives), and SAFE progress messages.

-   `SAFE_distribution`: Character or `NULL.` Simulation family to use for SAFE. Must match a `sim_family` value in the formulas table. If `NULL` and the chosen `effect_type` has a default simulation family (`default_safe_family == "yes"`), that default is used. If no default is defined, the function does not alter the filtered formulas.

-   `sigma_matrix`: List or `NULL.` Optional list of per-observation variance–covariance matrices used by SAFE (length equal to the number of rows implied by `...`). Each element should be a symmetric, positive definite matrix conformable with the simulated parameter vector required by the chosen effect formula. If `NULL`, internal defaults are used.
:::

### SAFE bootstrap length

We'll use this function to evaluate two bootstrap lengths: 100 and 1,000.

We'll run a Monte Carlo simulation 1,000 times just to make this easy. (See below for full 1e5 simulation results)

```{r, how_many_n_boots}
#| label: how_many_n_boots
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

# Create some scenarios based on combinations of sample size and boot-length
scenario <- CJ(true_mean1 = 13.4, true_sd1 = 4.6,
               true_mean2 = 16.1, true_sd2 = 3.9,
               sample_size1 = c(5, 10, 150))
scenario[, sample_size2 := sample_size1]
scenario

```

Now let's run this in a foreach loop. If you'd rather not run this on your machine, if `run` == FALSE, the code will just load the already finished simulation. Note that in this simulation, we're evaluating SAFE estimates using only the 'true' values, instead of simulated data.

```{r, how_many_n_boots_run}
#| label: how_many_n_boots_run
#| eval: false
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

out <- list()
sub_scenario <- c()
N <- 1000
res <- list()

clust_out <- prepare_cluster(n = N)

res <- foreach(i = 1:N, 
              .options.snow = clust_out$options,
             .errorhandling = "stop",
                  .packages = c("data.table", "MASS",
                           "crayon", "tmvtnorm")) %dopar% {

 # Now calculate SAFE with 100 bootstraps                                         
 out[[1]] <- eff_size(x1 = scenario$true_mean1,
                 x2 = scenario$true_mean2,
                 sd1 = scenario$true_sd1,
                 sd2 = scenario$true_sd2,
                 n1 = scenario$sample_size1,
                 n2 = scenario$sample_size2,
                 effect_type = "lnRoM",
                 SAFE = TRUE,
                 verbose = FALSE,
                 parallelize = FALSE,
                 SAFE_boots = 100) 
 
  out[[1]] <- data.table(out[[1]],
                         sample_size = scenario$sample_size1,
                         boots = 100)
  out[[1]]
  
  # Now with 1000 bootstraps
  out[[2]] <- eff_size(x1 = scenario$true_mean1,
                 x2 = scenario$true_mean2,
                 sd1 = scenario$true_sd1,
                 sd2 = scenario$true_sd2,
                 n1 = scenario$sample_size1,
                 n2 = scenario$sample_size2,
                 effect_type = "lnRoM",
                 SAFE = TRUE,
                 verbose = FALSE,
                 parallelize = FALSE,
                 SAFE_boots = 1000) 
  
  out[[2]] <-  data.table(out[[2]],
                         sample_size = scenario$sample_size1,
                         boots = 1000)
  out[[2]]   
  
  return(rbindlist(out))
}

stopCluster(clust_out$clust)
res.dt <- rbindlist(res)

saveRDS(res.dt, "builds/dummy_bootstrap_simulation.Rds")

```

```{r, read_bootstrap_sim}
#| label: read_bootstrap_sim
#| echo: false
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"
res.dt <- readRDS("builds/dummy_bootstrap_simulation.Rds")

```

Let's look at the dispersion of the SAFE estimates as a function of sample size and bootstrap length.

```{r, boots_results}
#| label: boots_results
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

head(res.dt)

# First, the point estimates:
ggplot(data = res.dt, 
       aes(x = as.factor(sample_size), y = yi_safe, 
           fill = boots,
           group = interaction(sample_size, boots)))+
  geom_jitter(alpha = .5, shape = 21,
              position = position_jitterdodge(dodge.width = 1))+
  geom_violin(position = position_dodge(width = 1))+
  ylab("SAFE Point Estimate")+
  xlab("Scenario sample size")+
  scale_fill_scico("Number of SAFE bootstraps",
                   palette = "hawaii")+
  theme_bw()+
  theme(panel.border = element_blank(),
        panel.grid = element_blank(),
        strip.background = element_blank(),
        legend.position = "bottom")

```

Now let's look at the sampling variance estimates:

```{r, boots_results_var}
#| label: boots_results_var
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

ggplot(data = res.dt, 
       aes(x = as.factor(sample_size), y = vi_safe, 
           fill = boots,
           group = interaction(sample_size, boots)))+
  geom_jitter(alpha = .5, shape = 21,
              position = position_jitterdodge(dodge.width = 1))+
  geom_violin(position = position_dodge(width = 1))+
  ylab("SAFE Variance Estimate")+
  xlab("Scenario sample size")+
  scale_fill_scico("Number of SAFE bootstraps",
                   palette = "hawaii")+
  theme_bw()+
  theme(panel.border = element_blank(),
        panel.grid = element_blank(),
        strip.background = element_blank(),
        legend.position = "bottom")


```

It looks like the dispersion in estimates is shaped by number of SAFE bootstraps and improves considerably with 1,000 bootstraps, especially at low sample sizes. Another way to visualize this (which we'll use below in the proper 1e5 simulations) is to plot the standard deviation.

```{r, boots_results_sd}
#| label: boots_results_sd
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

res.summary <- res.dt[, .(sd_point_estimate = sd(yi_safe),
                       sd_variance_estimate = sd(vi_safe)),
                   by = .(sample_size, boots)]
res.summary

# Let's melt this to make a single plot:
res.summary.mlt <- melt(res.summary,
                        id.vars = c("sample_size", "boots"))
res.summary.mlt
setorder(res.summary.mlt, boots)

ggplot(data = res.summary.mlt,
       aes(x = boots, y = value, color = sample_size,
           group = interaction(sample_size)))+
  geom_path()+
  geom_point(size = pt_size)+
  facet_wrap(~variable,
             scales = "free_y",
             labeller = as_labeller(c("sd_point_estimate" = "Point estimate",
                                      "sd_variance_estimate" = "Variance estimate")))+
  scale_color_scico("Scenario sample size",
                   palette = "hawaii")+
  xlab("SAFE bootstrap length")+
  ylab("Standard deviation of estimate")+
  theme_bw()+
  theme(panel.border = element_blank(),
        panel.grid = element_blank(),
        strip.background = element_blank(),
        legend.position = "bottom")

```

## Bootstrap scenario results

Now let's look at the full Monte Carlo simulation results for the influence of bootstrap length.

```{r, read_all_bootstrap_sim}
#| label: read_all_bootstrap_sim
#| echo: true
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

sim_results <- readRDS("builds/all_scenarios_summarized.Rds")

sim_results <- sim_results[calculation == "SD", ]

```

### Speed

Once again, here's our novel effect size 'speed', 1 / x. How do our SAFE estimates of speed and its sampling variance change as we changed the number of bootstraps?

```{r, speed_num_boots}
#| label: speed_num_boots
#| echo: false
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

y_lab <- "Standard deviation of estimate"

theme_SAFE <- theme_bw()+
  theme(panel.border = element_blank(),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        panel.grid = element_blank(),
        strip.background = element_blank())

pt_size <- 2

facet_labs <- as_labeller(c("point" = "Point estimate",
                            "variance" = "Variance estimate"))

sub_dat <- sim_results[effect_type == "reciprocal", ]

setorder(sub_dat, boots)

ggplot(data = sub_dat, 
            aes(x = boots, y = value, 
                color = sample_size,
                group = sample_size))+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(y_lab)+
  xlab("SAFE bootstrap length")+
  scale_x_log10(breaks = c(1e1, 1e2, 1e3, 1e4, 1e5, 1e6, 1e7))+
  scale_color_scico(name = "Scenario sample size", 
                    palette = "hawaii")+
  facet_wrap(~estimate_of, 
             scales = "free_y",
             labeller = facet_labs)+
  ggtitle("Speed (1 / x)")+
  theme_SAFE+
  theme(legend.position = "bottom")

```

### SMD 4-multivariate normal

```{r, SMD_mvnormal_num_boots}
#| label: SMD_mvnormal_num_boots
#| echo: false
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

sub_dat <- sim_results[effect_type == "SMD_normal", ]
setorder(sub_dat, boots)
# sub_dat

ggplot(data = sub_dat, 
            aes(x = boots, y = value, 
                color = sample_size1,
                group = sample_size1))+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(y_lab)+
  xlab("SAFE bootstrap length")+
  scale_x_log10(breaks = c(1e1, 1e2, 1e3, 1e4, 1e5, 1e6, 1e7))+
  scale_color_scico(name = "Scenario sample size", 
                    palette = "hawaii")+
  facet_wrap(~estimate_of, 
             scales = "free_y",
             labeller = facet_labs)+
  ggtitle("SMD (4-multivariate normal SAFE distribution)")+
  theme_SAFE+
  theme(legend.position = "bottom")

```

### SMD 2-multivariate normal and 2-Wishart

```{r, SMD_Wishart_num_boots}
#| label: SMD_Wishart_num_boots
#| echo: false
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

sub_dat <- sim_results[effect_type == "SMD_Wishart", ]
setorder(sub_dat, boots)

ggplot(data = sub_dat, 
            aes(x = boots, y = value, 
                color = sample_size1,
                group = sample_size1))+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(y_lab)+
  xlab("SAFE bootstrap length")+
  scale_x_log10(breaks = c(1e1, 1e2, 1e3, 1e4, 1e5, 1e6, 1e7))+
  scale_color_scico(name = "Scenario sample size", 
                    palette = "hawaii")+
  facet_wrap(~estimate_of, 
             scales = "free_y",
             labeller = facet_labs)+
  ggtitle("SMD (2-multivariate normal, 2-Wishart)")+
  theme_SAFE+
  theme(legend.position = "bottom")

```

### lnCVR 4-multivariate normal

```{r, lnCVR_mvnormal_num_boots}
#| label: lnCVR_mvnormal_num_boots
#| echo: false
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

sub_dat <- sim_results[effect_type == "lnCVR_normal", ]
setorder(sub_dat, boots)

ggplot(data = sub_dat, 
            aes(x = boots, y = value, 
                color = sample_size1,
                group = sample_size1))+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(y_lab)+
  xlab("SAFE bootstrap length")+
  scale_x_log10(breaks = c(1e1, 1e2, 1e3, 1e4, 1e5, 1e6, 1e7))+
  scale_color_scico(name = "Scenario sample size", 
                    palette = "hawaii")+
  facet_wrap(~estimate_of, 
             scales = "free_y",
             labeller = facet_labs)+
  ggtitle("lnCVR (4-multivariate normal)")+
  theme_SAFE+
  theme(legend.position = "bottom")

```

### lnCVR 2-multivariate normal and 2-Wishart

```{r, lnCVR_Wishart_num_boots}
#| label: lnCVR_Wishart_num_boots
#| echo: false
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

sub_dat <- sim_results[effect_type == "lnCVR_Wishart", ]
setorder(sub_dat, boots)

ggplot(data = sub_dat, 
            aes(x = boots, y = value, 
                color = sample_size1,
                group = sample_size1))+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(y_lab)+
  xlab("SAFE bootstrap length")+
  scale_x_log10(breaks = c(1e1, 1e2, 1e3, 1e4, 1e5, 1e6, 1e7))+
  scale_color_scico(name = "Scenario sample size", 
                    palette = "hawaii")+
  facet_wrap(~estimate_of, 
             scales = "free_y",
             labeller = facet_labs)+
  ggtitle("SMD (2-multivariate normal, 2-Wishart)")+
  theme_SAFE+
  theme(legend.position = "bottom")

```

### Log Odds Ratio (lnOR)

```{r, lnOR_num_boots}
#| label: lnOR_num_boots
#| echo: false
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

sub_dat <- sim_results[effect_type == "lnOR", ]
setorder(sub_dat, boots)

ggplot(data = sub_dat, 
            aes(x = boots, y = value, 
                color = n1,
                group = n1))+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(y_lab)+
  xlab("SAFE bootstrap length")+
  scale_x_log10(breaks = c(1e1, 1e2, 1e3, 1e4, 1e5, 1e6, 1e7))+
  scale_color_scico(name = "Scenario sample size", 
                    palette = "hawaii")+
  facet_wrap(~estimate_of, 
             scales = "free_y",
             labeller = facet_labs)+
  ggtitle("lnOR")+
  theme_SAFE+
  theme(legend.position = "bottom")

```

### Log Risk Ratio (lnRR)

```{r, lnRR_num_boots}
#| label: lnRR_num_boots
#| echo: false
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

sub_dat <- sim_results[effect_type == "lnRR", ]
setorder(sub_dat, boots)

ggplot(data = sub_dat, 
            aes(x = boots, y = value, 
                color = n1,
                group = n1))+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(y_lab)+
  xlab("SAFE bootstrap length")+
  scale_x_log10(breaks = c(1e1, 1e2, 1e3, 1e4, 1e5, 1e6, 1e7))+
  scale_color_scico(name = "Scenario sample size", 
                    palette = "hawaii")+
  facet_wrap(~estimate_of, 
             scales = "free_y",
             labeller = facet_labs)+
  ggtitle("lnRR")+
  theme_SAFE+
  theme(legend.position = "bottom")

```

### Hardy Weinberg Disequilibrium

```{r, lnHWE_A_num_boots}
#| label: lnHWE_A_num_boots
#| echo: false
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

sub_dat <- sim_results[effect_type == "lnHWE_A", ]
setorder(sub_dat, boots)

ggplot(data = sub_dat, 
            aes(x = boots, y = value, 
                color = n,
                group = n))+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(y_lab)+
  xlab("SAFE bootstrap length")+
  scale_x_log10(breaks = c(1e1, 1e2, 1e3, 1e4, 1e5, 1e6, 1e7))+
  scale_color_scico(name = "Scenario sample size", 
                    palette = "hawaii")+
  facet_wrap(~estimate_of, 
             scales = "free_y",
             labeller = facet_labs)+
  ggtitle("Hardy Weinburg Disequilibrium")+
  theme_SAFE+
  theme(legend.position = "bottom")

```

# Paired design Monte Carlo {#sec-paired-designs-monte-carlo}

Now let's evaluate how different values of $r$ affect bias and relative bias of SAFE and plugin estimators. We included an $r$ of 0, 0.5, and 0.8 and we ran each scenario 1e5 times. We did this for lnRoM and SMD and lnCVR—the 3 main effect sizes with paired designs.

### SMD
First, the bias of the point estimates:
```{r}
#| label: SMD_paired_point_bias
#| eval: true
#| echo: false
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

paired_dat <- readRDS("builds/paired_scenarios_bias.Rds")
paired_dat
sub_dat <- paired_dat[effect_type == "SMD", ]

unique(sub_dat$estimate)
labs = c("first_estimate" = "Cohen's d",
           "second_estimate" = "Hedges' g",
           "safe_estimate" = "SAFE")

pal = c("first_estimate" = "#A8DADC",
           "second_estimate" = "#1D3557",
           "safe_estimate" = "#E63946")
pt_size <- 2

unique(sub_dat$estimand)
unique(sub_dat$r)

facet_labs <- as_labeller(c("first_estimand" = "Cohen's d estimand",
                            "second_estimand" = "Hedges' g estimand",
                            "safe_estimand" = "SAFE estimand",
                            "0" = "r = 0",
                            "0.5" = "r = 0.5",
                            "0.8" = "r = 0.8"))
setorder(sub_dat, n)

p.point <- ggplot(data = sub_dat[measure %in% c("bias") &
                                   estimand == "first_estimand",],
                  aes(x = n, y = value, 
                      color = estimate,
                      group = estimate))+
  facet_wrap(~r, labeller = facet_labs)+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab("Bias of point estimate")+
  xlab("Sample size")+
  ggtitle("SMD")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs)+
  coord_cartesian(ylim = c(-0.1, 0.1))+
  ggtitle("SMD")+
  theme_SAFE+
  theme(legend.position = "bottom") 
p.point

```

And the relative bias of the variance estimates. Remember that these are in comparison to 3 potential estimands (Cohen's d, Hedges' g, and SAFE).
```{r}
#| label: SMD_paired_rel_bias
#| eval: true
#| echo: false
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

sub_dat$estimand <- factor(sub_dat$estimand,
                           levels = c("first_estimand", "second_estimand",
                                      "safe_estimand"))
p.variance <- ggplot(data = sub_dat[measure %in% c("rel_bias"), ], 
                     aes(x = n, y = value, color = estimate,
                         group = estimate))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab("Relative bias of variance estimates (%)")+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = c(plugin_1st = "Cohen's d", 
                                plugin_2nd = "Hedges' g", 
                                safe = "SAFE"))+
  # coord_cartesian(ylim = c(-50, 50))+
  facet_grid(estimand ~ r, 
             labeller = facet_labs)+
  theme_SAFE+
  theme(legend.position = "bottom")

p.variance
```

### lnRoM
First, the bias of the point estimates:
```{r}
#| label: lnRoM_paired_point_bias
#| eval: true
#| echo: false
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

sub_dat <- paired_dat[effect_type == "lnRoM" & !is.na(value), ]
sub_dat <- droplevels(sub_dat)

unique(sub_dat$variable)
unique(sub_dat$estimate)

labs = c("first_estimate" = "1st order plugin",
           "second_estimate" = "2nd order plugin",
           "safe_estimate" = "SAFE")

pal = c("first_estimate" = "#A8DADC",
           "second_estimate" = "#1D3557",
           "safe_estimate" = "#E63946")
pt_size <- 2

unique(sub_dat$estimand)
facet_labs <- as_labeller(c("first_estimand" = "1st order estimand",
                            "second_estimand" = "2nd order plugin estimand",
                            "safe_estimand" = "SAFE estimand",
                            "0" = "r = 0",
                            "0.5" = "r = 0.5",
                            "0.8" = "r = 0.8"))
pt_size <- 2

unique(sub_dat$estimand)
unique(sub_dat$r)

setorder(sub_dat, n)

p.point <- ggplot(data = sub_dat[measure %in% c("bias") &
                                   estimand == "first_estimand",],
                  aes(x = n, y = value, 
                      color = estimate,
                      group = estimate))+
  facet_wrap(~r, labeller = facet_labs)+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab("Bias of point estimates")+
  xlab("Sample size")+
  ggtitle("SMD")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs,
                     drop = TRUE)+
  coord_cartesian(ylim = c(-0.01, 0.01))+
  ggtitle("lnRoM")+
  theme_SAFE+
  theme(legend.position = "bottom") 
p.point

```

And the relative bias of the variance estimates. 
```{r}
#| label: SMD_paired_rel_bias
#| eval: true
#| echo: false
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"


p.variance <- ggplot(data = sub_dat[measure %in% c("rel_bias"), ], 
                     aes(x = n, y = value, color = estimate,
                         group = estimate))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab("Relative bias of variance estimates")+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = c(first_estimate = "1st order plugin", 
                                plugin_2nd = "2nd order plugin", 
                                safe = "SAFE"))+
  # coord_cartesian(ylim = c(-50, 50))+
  facet_grid(r~estimand, 
             labeller = facet_labs)+
  theme_bw()+
  # theme_SAFE+
  theme(legend.position = "bottom")
p.variance

```

### lnCVR
First, the bias of the point estimates:
```{r}
#| label: SMD_paired_point_bias
#| eval: true
#| echo: false
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

sub_dat <- paired_dat[effect_type == "lnCVR" & !is.na(value), ]
sub_dat <- droplevels(sub_dat)

unique(sub_dat$variable)
unique(sub_dat$estimate)

labs = c("first_estimate" = "1st order plugin",
           "second_estimate" = "2nd order plugin",
           "safe_estimate" = "SAFE")

pal = c("first_estimate" = "#A8DADC",
           "second_estimate" = "#1D3557",
           "safe_estimate" = "#E63946")
pt_size <- 2

unique(sub_dat$estimand)
facet_labs <- as_labeller(c("first_estimand" = "1st order estimand",
                            "second_estimand" = "2nd order plugin estimand",
                            "safe_estimand" = "SAFE estimand",
                            "0" = "r = 0",
                            "0.5" = "r = 0.5",
                            "0.8" = "r = 0.8"))
pt_size <- 2

unique(sub_dat$estimand)
unique(sub_dat$r)

setorder(sub_dat, n)

p.point <- ggplot(data = sub_dat[measure %in% c("bias") &
                                   estimand == "first_estimand",],
                  aes(x = n, y = value, 
                      color = estimate,
                      group = estimate))+
  facet_wrap(~r, labeller = facet_labs)+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(point_lab)+
  xlab("Sample size")+
  ggtitle("SMD")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = labs,
                     drop = TRUE)+
  coord_cartesian(ylim = c(-0.025, 0.025))+
  ggtitle("SMD")+
  theme_SAFE+
  theme(legend.position = "bottom") 
p.point

```

And the relative bias of the variance estimates. Remember that these are in comparison to 3 potential estimands (Cohen's d, Hedges' g, and SAFE).
```{r}
#| label: SMD_paired_rel_bias
#| eval: true
#| echo: false
#| attr-source: "style='font-size: 0.8em;'"
#| attr-output: "style='font-size: 0.7em'"

p.variance <- ggplot(data = sub_dat[measure %in% c("rel_bias"), ], 
                     aes(x = n, y = value, color = estimate,
                         group = estimate))+
  geom_hline(yintercept = 0)+
  geom_path(lwd = 1)+
  geom_point(size = pt_size)+
  ylab(var_lab)+
  xlab("Sample size")+
  scale_color_manual(name = "Estimator", 
                     values = pal,
                     labels = c("first_estimate" = "1st order plugin", 
                                "second_estimate" = "2nd order plugin", 
                                "safe_estimate" = "SAFE"))+
  # coord_cartesian(ylim = c(-50, 50))+
  facet_grid(r~estimand, 
             labeller = facet_labs)+
  theme_SAFE+
  theme(legend.position = "bottom")
p.variance

```

# Conclusion

SAFE bootstrapping is a remarkably simple and effective way to calculate bias-corrected point estimates and sampling variances for meta-analyses. Based on our simulations, this method performs similarly to plugin estimates and requires only 1e5 (though 1e6 is marginally better) bootstraps. SAFE allows researchers to calculate sampling variance for custom effect sizes with only minimal statistical knowledge (and without spooky Calculus).
